Lattice Boltzmann simulations of magnetically responsive\\ anisotropic particle stabilized emulsions

### 2025-03-05 chapters/paper2.tex
Changes 1. Reworded introduction and lit review and addressed grammar and vocab issues 2. Fixed some referencing problems in the document 3. Fleshed out introduction in paper 2 4. Added sections for results in paper 3

@@

We characterize the microstructure changes observed through calculating
a characteristic length scale, called the average domain size. The
average domain size is calculated using the first moment of the 3D
averaged structure factor. Past work has shown that this method
correctly reproduces the scaling expected in phase separating systems.
The structure factor is calculated from the fourier transform of the
fluctuations in the order parameter. \cite{kendon_inertial_2001}

\begin{equation}
S(\mathbf{k}, t) = \widetilde{\phi}(\mathbf{k}, t)\widetilde{\phi}(\mathbf{-k}, t)
\end{equation}

\(\mathbf{k}\) represents the wave vector in each direction of each
lattice grid point. The fluctuations in the order parameter are
calculated from the order parameter of the bijel defined as
\(\widetilde{\phi} = \phi - \langle \phi \rangle\). \(\phi\) is
calculated through subtraction of the red from the blue fluid density,
\(\phi = \rho^b - \rho^r\). The average domain size is commonly
calculated using the spherically averaged structure factor where all
points corresponding to \(k = |\mathbf{k}|\) are binned, where \(k\) is
the magnitude of \(\mathbf{k}\).

\begin{equation}
S(k, t) = \frac{1}{n} \sum_{k - \Delta < k < k + \Delta} S(\mathbf{k}, t)
\label{eq:structure_factor_calc}
\end{equation}

The average domain size can then be calculated from the spherically
averaged structure factor after calculating the first moment of the
structure factor using the following expression.

\begin{equation}
L(t) = 2\pi \frac{\sum S(k, t) }{ \sum k S(k, t) }
\end{equation}

We found that this method for determining domain size generally
overestimated the observed channel diameter. However, previous studies
have demonstrated that the scaling behavior of the first-moment-derived
domain size aligns with the expected phase separation in spinodal
decomposition. The overestimation arises due to the limited number of
samples at small k values, where the bin size is roughly equivalent to
the k value. Additionally, in simulations with periodic domains, both
the structure factor and computed length scale are influenced by
finite-size effects. Nonetheless, this measurement still provides
valuable insight into trends in the average domain size.

To correlate the microstructure evolution to the ordering of the
particle monolayer to the magnetic field, we calculate the nematic order
parameter. \cite{cuesta_monte_1990} First, the average orientational
tensor is calculated from the particle orientations in Eq
\ref{eq:orientational_order_tensor}

\begin{equation}
Q = \frac{1}{n_p} \sum_{i} (\frac{3}{2}[\hat{\mathbf{o}}_i \otimes \hat{\mathbf{o}}_i] - \frac{1}{2}\mathbf{I})
\label{eq:orientational_order_tensor}
\end{equation}

In Eq \ref{eq:orientational_order_tensor}, \(\hat{\mathbf{o}}_i\)
defines the orientation vector of particle \(i\) and \(\mathbf{I}\) is
the identity matrix. The nematic director or average particle
orientation is the eigenvector corresponding to the largest eigenvalue
of \(Q\) which is the value of the nematic order parameter, \(S\). The
assembly of particles can be said to have nematic order if
\(S \geq 0.3\) and is fully aligned to the director if \(S = 1.0\).
\cite{cuesta_monte_1990}

@@

Precise characterization of the anisotropic microstructure is essential
for the tailored design of application-specific porous materials. To
quantify the characteristic length scales in each Cartesian direction,
we compute the second-moment-derived structure factor without radial
averaging. \cite{jansen_bijels_2011} Following the structure factor
calculation described in Eq. \ref{eq:structure_factor_calc}, the second
moment of the structure factor in direction \(\beta\) is determined for
each Cartesian coordinate axis as

\begin{equation}
L_{\beta}(t) = 2 \pi \sqrt{\frac{\sum_{\mathbf{k_{\beta}}}  S(\mathbf{k}, t) }{\sum_{\mathbf{k_{\beta}}}  k_{\beta}^2S(\mathbf{k}, t)}}
\label{eq:second_moment_domain}
\end{equation}

From Eq \ref{eq:second_moment_domain}, we can calculate the length scale
in each cartesian direction, \(L_x, L_y, L_z\). Here, we make a further
refinement simplifying the domain sizes to directions parallel and
orthogonal to the magnetic field. These result in the length scales and
their associated calculation, \(L_{\parallel} = L_{z}\) and
\(L_{\perp} = \frac{L_x + L_y}{2}\). Tortuosity has been defined using
the geometric, diffusional and hydraulic tortuosities. Here we define
the diffusional tortuosity, \(\tau = \frac{D}{D_{eff}}\) which
characterizes the ratio between the free diffusion of a tracer and the
effective diffusion of a tracer in the material.

To calculate the tortuosities of the structures generated from our
simulations, we first calculate \(\phi\) of the system over time from
the filled densities. Many bijels are synthesized from monomer diffused
into the oil phase, or the oil phase itself is intrinsically
polymerizable. Our next step is to binarize the order parameter with the
condition, \(\phi \leq 0 = 0\) and \(\phi > 0 = 1\) which reflects a
solid red fluid phase and void blue fluid. We then mask all unconnected
components, using the \texttt{label} function in \texttt{skimage}
leaving only the percolating domain representing the bijel.
\cite{van2014scikit} We utilize the python package taufactor which
solves the diffusion equation for the structure, calculating the time
taken for a tracer to go from one side of the system to the other.
\cite{kench_taufactor_2023} We utilize the periodic boundary condition
solver for this investigation. In previous work, we identified that the
length scale derived from the second moment of the structure factor is
only useful for identifying length scale anisotropy. Therefore, we plot
the directional second moment derived length scales and tortuosity in
Figure \ref{fig:domain_size_aniso-field_on}.

@@

In Figure \ref{fig:particle_viz-field_on}, we see the effect of the
application of the magnetic field on the monolayer. The initial
configuration of particles at the monolayer show that the orientations
are random and that the particles tend to lie flat on the interface.
However, upon application of the magnetic field the particles orient to
the field direction. The effect that Gunther. et al.~observed for bijels
stabilized with anisotropic particles can also be observed in these
snapshots as the particles can be observed to have more local order in
the final state than in the initial state with no applied field. We
observe that for bijels stabilized by prolate ellipsoids, we see the
adoption of more regular arrangements of the colloids on the interface.
We investigate the effect of local particle ordering next. Past work by
Bresme and Faraudo and Davies investigating the reorientation of
particles at interfaces under magnetic fields have shown that a critical
magnetic field exists, above which the particle orients to the field and
below which the particle orientation is a function of the capillary
force and magnetic field derived force. \cite{davies_interface_2014}
\cite{bresme_orientational_2007} Particle adsorption into the interface
with microparticles have also been shown to be irreversible due to the
high adsorption energy. Upon particle reorientation, the microstructure
change can be driven solely through particle unjamming and re-jamming,
or it can be driven through the particles dragging the interface during
this process. The mechanism of unjamming and re-jamming can be
investigated using the average interfacial angle, \(\psi\).

The interface angle of each particle is defined as
\(\psi = \arccos{\frac{ \hat{\mathbf{o_i}}\cdot \hat{\mathbf{n_i}}}{|\hat{\mathbf{o_i}}| |\hat{\mathbf{n_i}}|}}\)
where \(\hat{\mathbf{n_i}}\) is the normal of the interface closest to
the center of the particle \(i\) and \(\hat{\mathbf{o_i}}\) is the
orientation of particle \(i\) which follows the symmetry axis of the
ellipsoid. For prolate particles, they prefer to lie flat on the
interface, meaning that with no field \(\psi = 90^{\circ}\).
\(\hat{\mathbf{n_i}}\) is calculated by first using the particle filling
algorithm described and calculating the order parameter. Next, a mesh of
the isocontour at \(\phi = 0\) is generated using the marching cubes
algorithm in scikit-image with a box length of 4. \cite{van2014scikit}
Next, the vertex closest to the center of particle \(i\) was found.
\(\hat{\mathbf{n_i}}\) was then obtained from the normal corresponding
to the vertex identified in the previous step.

Another factor affected by the application of the magnetic field is the
Steinhardt 6 fold order parameter, \(Q_6\), which characterized the
local particle ordering. \cite{steinhardt_bond-orientational_1983} It
has been used to characterize crystallization, glass transitions and
crystal structures for colloidal and molecular crystals. We utilize the
average across all particles on the interface, \(\langle Q_6 \rangle\)
and track this parameter as a function of time. Past work has identified
that local crystallization of colloidal crystals takes place at
\(\langle Q_6 \rangle = 0.38\), and that
\(\langle Q_6 \rangle^{HCP} = 0.485\).
\cite{steinhardt_bond-orientational_1983} \cite{toxvaerd_role_2020}
\cite{mickel_shortcomings_2013} We utilize a voronoi cell derived
definition of the neighbor list of each particle, addressing some of the
shortcomings highlighted in the original distance based definition.
\cite{steinhardt_bond-orientational_1983}
\cite{mickel_shortcomings_2013} We first define a complex orientational
vector, \(q_{lm}\)

\begin{equation}
q_{lm}(i) = \frac{1}{N_b(i)} \Sigma_{j = 1}^{N_b(i)} Y_{lm}(\vec{R}_{ij})
\end{equation}

Where \(N_b\) is the number of nearest neighbors of particle \(i\),
\(l\) is controls to the degree of spherical harmonic to be calculated,
\(m\) is an integer defined as \(-l \leq m \leq l\) that defines which
spherical harmonic is being calculated and \(\vec{r}_{ij}\) is the
distance between two particle centers. The neighbor list \(N_b\) is
usually defined by a cutoff distance from the center of particle \(i\)
to its nearest neighbors \(j\). Steinhardt then defined a bond order
parameter \(q_l\) from \(q_{lm}\) as

\begin{equation}
Q_{l}(i) = \sqrt{\frac{4 \pi}{2l + 1} \Sigma_{m = -l}^{l} |q_{lm}(i)|^2}
\end{equation}

>>>>

Stimuli responsive emulsions have seen interest for use in fields such
as enhanced oil recovery, porous material templates, in-situ reaction
seperation systems and drug delivery scaffolds.
\cite{tang_stimuli-responsive_2016,} \cite{bago_rodriguez_capsules_2019}
These applications are enabled through stimuli response driven
microstructure modification, changing the permeability, sueface area and
morphology of the emulsion. Leveraging advances in particle synthesis
techniques and citing concerns over surfactant toxicity and
environmental concerns, particle stabilized emulsions have made a
resurgence in popularity after their discovery in 1907 by Ramsden and
Pickering. \cite{rozynek_opening_2019} \cite{zhou_magnetic_2011} Stimuli
response has been enabled in particle stabilized emulsions using pH,
temperature or chemical composition changes.
\cite{maingret_dextran-based_2020} \cite{lin_ph-responsive_2018}
\cite{raju_ph-responsive_2018} \cite{zhang_ph_2022}

Owing to its tortuous, co-continuous domain morphology, bicontinuous
interfacially jammed emulsion gels (bijel) are studied for their
possible applications to fabricate membranes, catalyst supports, battery
electrodes and pharmaceutical products. \cite{yabuno_preparation_2020}
\cite{samdani_bicontinuous_2017} \cite{cha_bicontinuous_2019}
\cite{garcia_scalable_2019} \cite{santiago_cordoba_aerobijels_2020}
Bijels are formed through the arresting of spinodal decomposition of
partially miscible fluids. \cite{stratford_colloidal_2005}
\cite{herzig_bicontinuous_2007} \cite{tavacoli_novel_2011} Once phase
separation begins, the incipient interface begins to sweep through the
system causing particles to begin to adsorbing into it. Once the
interfacial area matches that of the cross sectional area of the
adsorbed particles, the structure jams in place.
\cite{stratford_colloidal_2005} \cite{herzig_bicontinuous_2007}
\cite{tavacoli_novel_2011} After their discovery in simulations and
subsequent synthesis in experiments, a rich diversity of bijel
literature has developed around various fluid systems that allow for
bijel synthesis, methods to fabricate them at scale and their
applications.

Bijels stabilized by ellipsoidal particles exhibit distinct domain evolution timescales compared to those stabilized by 
spherical particles \cite{gunther_timescales_2014}. Unlike spherical particles, ellipsoidal particle reorientation at the 
interface influences microstructure. After jamming, additional timescales emerge due to capillary interactions that drive 
particle reorientation and prolonged domain coarsening \cite{gunther_timescales_2014}. Bijels behave as 2D glasses percolating 
in 3D space, with locally jammed particles as observed by Günther et al. 
\cite{ching_bijel_2022, torquato_jammed_2010, gunther_timescales_2014}. Studies on jammed suspensions indicate subdiffusive 
particle movement, further confirming local jamming in bijel monolayers \cite{savelev_diffusion_2006}.

Magnetic fields have gained interest as external stimuli due to their targeted interactions, biocompatibility, and material 
versatility. The structural response of magnetically responsive hydroxyapatite crystals depends on initial particle orientation
\cite{nakayama_stimuli-responsive_2018}. Magnetophoresis, where particles move along field lines, has been observed in Pickering 
emulsions stabilized by magnetic microspheres \cite{tham_magnetophoresis_2021}. Structural deformation and de-emulsification occur due 
to particle migration or tilting at the interface \cite{yang_rapid_2020, misra_magnetic_2020}. However, how these effects influence 
bijel microstructure post-jamming remains unclear. Given the locally jammed nature of bijels, magnetic field responses, combined with 
anisotropic particle stabilization, may offer a means to control bijel microstructure.

This work investigates the structural response of bijels stabilized by anisotropic magnetic particles. Using a Lattice Boltzmann Method 
coupled with Molecular Dynamics, we simulate a binary fluid system where ellipsoidal particles, modeled as magnetic dipoles along their 
symmetry axis, stabilize bijel templates. Simulations without an applied field reveal anisotropic bijel formation, with microstructure 
evolution dependent on particle reordering at the interface, quantified using the six-fold Steinhardt order parameter \(\langle Q_6 \rangle\).

We further assess how initial microstructure influences structural changes, showing that the difference in order between initial and final 
particle monolayers dictates the extent of microstructure evolution. Additionally, applying and removing a magnetic field results in distinct 
microstructural properties and timescales, highlighting the importance of processing history. These findings demonstrate how magnetic fields, 
combined with anisotropic particle-stabilized bijels, enable on-demand microstructure switching for potential applications.

@@

To correlate the microstructure evolution to the ordering of the
particle monolayer to the magnetic field, we calculate the nematic order
parameter. \cite{veerman_phase_1992}

@@

Precise characterization of the anisotropic microstructure is essential
for the tailored design of application-specific porous materials. To
quantify the characteristic length scales in each Cartesian direction,
we compute the second-moment-derived structure factor without radial
averaging. \cite{jansen_bijels_2011} We also calculate the tortuosity 
in each cartesian direction. Here, we make a further
refinement simplifying the domain sizes to directions parallel and
orthogonal to the magnetic field. These result in the length scales and
their associated calculation, \(L_{\parallel} = L_{z}\) and
\(L_{\perp} = \frac{L_x + L_y}{2}\). We perform the same simplification
with the directional tortuosities.

Tortuosity has been defined using
the geometric, diffusional and hydraulic tortuosities. Here we define
the diffusional tortuosity, \(\tau = \frac{D}{D_{eff}}\) which
characterizes the ratio between the free diffusion of a tracer and the
effective diffusion of a tracer in the material. Therefore, we plot
the directional second moment derived length scales and tortuosity in
Figure \ref{fig:domain_size_aniso-field_on}.

@@

In Figure \ref{fig:particle_viz-field_on}, we see the effect of the
application of the magnetic field on the monolayer. The initial
configuration of particles at the monolayer show that the orientations
are random and that the particles tend to lie flat on the interface.
However, upon application of the magnetic field the particles orient to
the field direction. The effect that Gunther. et al.~observed for bijels
stabilized with anisotropic particles can also be observed in these
snapshots as the particles can be observed to have more local order in
the final state than in the initial state with no applied field. We
observe that for bijels stabilized by prolate ellipsoids, we see the
adoption of more regular arrangements of the colloids on the interface.
We investigate the effect of local particle ordering next. Past work by
Bresme and Faraudo and Davies investigating the reorientation of
particles at interfaces under magnetic fields have shown that a critical
magnetic field exists, above which the particle orients to the field and
below which the particle orientation is a function of the capillary
force and magnetic field derived force. \cite{davies_interface_2014}
\cite{bresme_orientational_2007} Particle adsorption into the interface
with microparticles have also been shown to be irreversible due to the
high adsorption energy. Upon particle reorientation, the microstructure
change can be driven solely through particle unjamming and re-jamming,
or it can be driven through the particles dragging the interface during
this process. The mechanism of unjamming and re-jamming can be
investigated using the average interfacial angle, $\langle \psi \rangle$
and the local ordering of the particles characterized with 
$\langle Q_6 \rangle$ Past work has identified that local crystallization 
of colloidal crystals takes place at \(\langle Q_6 \rangle = 0.38\). 
\cite{toxvaerd_role_2020}
